---
title: "Formal Exploratory Analysis"
author: "Anand Rajan"
date: "4/16/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(nlme)
library(lme4)
library(haven)
```

## Data Cleaning

```{r merging data, echo=FALSE, warning=FALSE}
midus_1 <- read_sav("./data/M2_P1_SURVEY_N4963_20200720.sav") %>% 
  select(M2ID,M2FAMNUM,B1PAGE_M2,B1PRSEX,B1PF7A,B1SA62A,B1SA62B,B1SA62C,B1SA62D,B1SA62E,B1SA62F,B1SA62G,B1SA62H,B1SA62I,B1SA62J,B1PTSEI,B1PB1)

midus_3 <- read_sav("./data/M2_P3_BTACT_N4512_20211123.sav") %>% 
         select(M2ID,B3TCOMPZ3,B3TEMZ3,B3TEFZ3) %>% 
        na.omit()

project4 <- load(file='./data/project4.rda')

midus_4 <-
  da29282.0001 %>% 
  select(M2ID,B4QCT_EA,B4QCT_EN,B4QCT_MD,B4QCT_PA,B4QCT_SA,B4QCT_PN) %>% 
  mutate(
    threat = B4QCT_EA + B4QCT_PA + B4QCT_SA,
    deprivation = B4QCT_EN + B4QCT_MD + B4QCT_PN
  ) %>% 
  select(M2ID, threat,deprivation)

milwaukee <- read_tsv("./data/MKE Data.tsv") %>% 
  select(M2ID,BACRAGE,BACRSEX,BACB1,BACF7A,BACAS62A,BACAS62B,BACAS62C,BACAS62D,BACAS62E,BACAS62F,BACAS62G,BACAS62H,BACAS62I,BACAS62J)

midus_df1 <- dplyr::inner_join(midus_4,midus_3)
midus_df2 <-dplyr::left_join(midus_df1,midus_1)
midus_df <- dplyr::left_join(midus_df2,milwaukee)
```

```{r}
midus <-
midus_df %>% 
  mutate(
    age = coalesce(BACRAGE,B1PAGE_M2),
    sex = coalesce(B1PRSEX,BACRSEX),
    race=coalesce(B1PF7A,BACF7A),
    education = coalesce(B1PB1,BACB1),
    tranquilizer = coalesce(B1SA62A,BACAS62A),
    stimulants = coalesce(B1SA62B,BACAS62B),
    painkillers = coalesce(B1SA62C,BACAS62C),
    depressants = coalesce(B1SA62D,BACAS62D),
    inhallants = coalesce(B1SA62E,BACAS62E),
    marijuana = coalesce(B1SA62F,BACAS62F),
    cocaine = coalesce(B1SA62G,BACAS62G),
    hallucinogens = coalesce(B1SA62H,BACAS62H),
    heroin = coalesce(B1SA62J,BACAS62J)
  ) %>% 
  rename(
    composite_cognition = B3TCOMPZ3,
    episodic_memory = B3TEMZ3,
    executive_func = B3TEFZ3,
    ses_index = B1PTSEI,
    family_id = M2FAMNUM
  ) %>% 
  filter(
    race %in% (1:6)
  ) %>% 
  mutate(
    race = as.factor(race),
    sex = as.factor(sex),
    education = as.factor(education),
    tranquilizer = as.factor(tranquilizer),
    stimulants = as.factor(stimulants),
    painkillers = as.factor(painkillers),
    depressants = as.factor(depressants),
    inhallants = as.factor(inhallants),
    marijuana = as.factor(marijuana),
    cocaine = as.factor(hallucinogens),
    hallucinogens = as.factor(hallucinogens),
    heroin = as.factor(heroin)
  ) %>% 
  mutate(
    tranquilizer = recode(tranquilizer,
                           '1' = "Yes",
                           '2' = "No"),
    stimulants = recode(stimulants,
                           '1' = "Yes",
                           '2' = "No"),
    painkillers = recode(painkillers,
                           '1' = "Yes",
                           '2' = "No"),
    depressants = recode(depressants,
                           '1' = "Yes",
                           '2' = "No"),
    inhallants = recode(inhallants,
                           '1' = "Yes",
                           '2' = "No"),
    marijuana = recode(marijuana,
                           '1' = "Yes",
                           '2' = "No"),
    cocaine = recode(cocaine,
                           '1' = "Yes",
                           '2' = "No"),
    hallucinogens = recode(hallucinogens,
                           '1' = "Yes",
                           '2' = "No"),
    heroin = recode(heroin,
                           '1' = "Yes",
                           '2' = "No"),
    race = recode(race,
                            '1' = "White",
                            '2' = "Black",
                            '3' = "Native American",
                            '4' = "Asian",
                            '5' = "Native Hawaiian or Pacific Islander",
                            '6'= "Other"
                  ),
    sex= recode(sex,
                '1'="Male",
                '2'="Female"),
    education = recode(education,
                       '1'='2')
  ) %>% 
  mutate(
    drug_use = case_when(tranquilizer == 'Yes'| stimulants == 'Yes'| painkillers == 'Yes'| depressants == 'Yes' | inhallants == 'Yes' | marijuana == 'Yes' | cocaine == 'Yes' | heroin == 'Yes'  ~ 'Yes',
                         TRUE ~ 'No')
  ) %>% 
  select(family_id,age,sex,race,education,tranquilizer,stimulants,painkillers,depressants,inhallants,marijuana,cocaine,hallucinogens,heroin,composite_cognition,episodic_memory,executive_func,threat,deprivation, drug_use) %>% 
  drop_na(threat,deprivation) %>% 
  janitor::clean_names()



midus <- within(midus,tranquilizer <- relevel(tranquilizer,ref="No"))
midus <- within(midus,stimulants <- relevel(stimulants,ref="No"))
midus <- within(midus,painkillers <- relevel(painkillers,ref="No"))
midus <- within(midus,inhallants <- relevel(inhallants,ref="No"))
midus <- within(midus,marijuana <- relevel(marijuana,ref="No"))
midus <- within(midus,cocaine <- relevel(cocaine,ref="No"))
midus <- within(midus,hallucinogens <- relevel(hallucinogens,ref="No"))
midus <- within(midus,depressants <- relevel(depressants,ref="No"))

```


After completion of data cleaning there were 1108 observations with 20 variables. A fairly substantial data set to run analyses on. Now lets dive into some exploratory analyses starting with looking at our outcome variable cognition.

# Exploratory 

To start, we will be looking out our outcome variable cognition. We can evaluate cognition through three different variables. We have composite cognition z scores, then cognition z scores for episodic memory and executive functioning. We will look at the distributions of all three outcome variables to potentially inform us of what sort of analysese we can run. Moreover we will look at univariate statistics for each of the cognition scores. 

```{r}
shapiro.test(midus$composite_cognition)
cognition_distribution <- ggplot(midus,aes(x=composite_cognition)) + 
                      geom_histogram(color = "blue")

cognition_distribution

```

The histogram indicates a fairly normal distribution of composite cognition z scores. Furthermore normality was tested via the shapiro-wilk test, and at the 0.05 significance level we conclude the distribution of scores is approximately normal.


```{r}
midus %>% 
  summarize(
    mean_composite = mean(composite_cognition),
    median_composite=median(composite_cognition),
    sd = sd(composite_cognition),
    min = min(composite_cognition),
    max = max(composite_cognition)
  ) %>% 
  knitr::kable()
```

The table above provides univariate statistics regarding composite cognition scores in the data set. 

Let us now look at episodic memory. 

```{r}
episodic_memory_distribution <-ggplot(midus,aes(x=episodic_memory)) + geom_histogram(color="blue")
shapiro.test(midus$episodic_memory)


episodic_memory_distribution
```


The histogram indicates the distribution of episodic memory z scores is right skewed. Normality was tested via the shapiro-wilk test, and at the 0.05 significance level, we conclude the distribution of scores is NOT normal. 

```{r}
midus %>% 
  summarize(
    mean_episodic = mean(episodic_memory),
    median_episodic=median(episodic_memory),
    sd = sd(episodic_memory),
    min = min(episodic_memory),
    max = max(episodic_memory)
  ) %>% 
  knitr::kable()
```

The table above provides univariate statistics on episodic memory z scores.

Let's now take a look at executive functioning.

```{r}
exec_functioning_distribution <- ggplot(midus,aes(x=executive_func)) + geom_histogram(color = "blue")
shapiro.test(midus$executive_func)

exec_functioning_distribution

```


The histogram indicates slight skewness, but mainly the existence of possible outliers. The shapiro wilks test also indicates non-normality, but I suspect this is due to outliers. 

```{r}
midus %>% 
  summarize(
    mean_exec = mean(executive_func),
    median_exec=median(executive_func),
    sd = sd(executive_func),
    min = min(executive_func),
    max = max(executive_func)
  ) %>% 
  knitr::kable()
```

The table above provides univariate statistics on executive function z-scores.



## Childhood Trauma

Childhood Trauma is broken down into 6 components scores. These components are emotional abuse,emotional neglect, minimization/denial,physical abuse, physical neglect, and sexual abuse. For this study, we will be categorizing trauma into two strata, Threat and Deprivation. Threat is the sum of the scores for emotional abuse, physical abuse and sexual abuse. While deprivation is the sum of emotional neglect, minimization/denial, and parental neglect. As you can see the division of the two strata make sense as one threat in the purview of childhood trauma refers to abuse, while deprivation refers to neglect. Now lets dive into each category. 

```{r}
midus %>% 
  summarize(
    mean_threat = mean(threat),
    median_threat=median(threat),
    sd = sd(threat),
    min = min(threat),
    max = max(threat)
  ) %>% 
  knitr::kable()

```

The table above provides univariate statistics for childhood threat scores. 

```{r}
midus %>% 
  summarize(
    mean_deprivation = mean(deprivation),
    median_deprivaton=median(deprivation),
    sd = sd(deprivation),
    min = min(deprivation),
    max = max(deprivation)
  )
```
 The table above provides univariate statistics regarding deprivation scores. 
 
 
```{r}
ggplot(midus,aes(x=threat)) +geom_histogram()
ggplot(midus,aes(x=deprivation)) +geom_histogram()
```


## Unadjusted Analysis

Now that we have looked at both cognition and childhood trauma individually as variables, lets explore the relationship between these two variables. 

### Composite Cognition

```{r}
par(mfrow=c(1,2))
plot(midus$threat,midus$composite_cognition)
plot(midus$deprivation,midus$composite_cognition)
```
The graphs above do not indicate any significant li


```{r}
fit1 <- lmer(composite_cognition ~ threat + deprivation + (1|family_id), data=midus)

summary(fit1)
```


### Episodic Memory

```{r}
par(mfrow=c(1,2))
plot(midus$threat,midus$episodic_memory)
plot(midus$deprivation,midus$episodic_memory)
```

```{r}
fit2 <- lmer(episodic_memory ~ threat + +deprivation + (1|family_id), data=midus)

summary(fit2)
```

### Executive Functioning

```{r}
par(mfrow=c(1,2))
plot(midus$threat,midus$executive_func)
plot(midus$deprivation,midus$executive_func)
```


```{r}
fit3 <- lmer(executive_func ~ threat + +deprivation + (1|family_id), data=midus)

summary(fit3)
```
# Evaluating Confounding


Now we will be getting into the meat of our analysis. First it should noted that the observations are in fact non-independent. Why is this the case? Well observations were not randomly sampled. We see sampling of participants from the same family, hence the inclusion of family ID in our data set. Cluster sampling was utilized. Therefore participants from the same family likely have similar levels of childhood trauma experience and similar cognition. Now given this, much of the analysis we could question the validity of the analysis run prior, but the purpose of the analysis before is to get an initial look at the data we are working with. the linear models fitted were merely to detect if there is any un-adjusted association.


Now we will delve into the real analysis. First lets start by looking at confounding. The variables we will be looking at is race, education, age, and sex. To evaluate confounding we will build linear mixed models evaluating the covariates association to both the outcome(composite_cognition,episodic_memory,executive_func)  


```{r}
boxplot(composite_cognition ~ race, data=midus)
boxplot(episodic_memory ~ race, data=midus)
boxplot(executive_func ~ race, data=midus)

race_lme <- lmer(composite_cognition~ race + (1|family_id), data=midus)
summary(race_lme)
```
To evaluate confounding we first fit a boxplot. From the box plot we see median composite cognition score is different across races. We see that the highest composite cognition score is among those who identified as native american, the lowest median composite cognition score is among those who identified as black. To further to see if there is an association between race and composite cognition, we it a linear model. To note, the observations are not considered independent, but I just want to see if there is an association while disregarding grouping. . To evaluate the specific parameters, lets introduce a mixed effects model to have greater interpretability. To begin, the mixed effect model tells us that 54.35% of the variance is not explained by the fixed effects. This indicates the importance of including grouping by family as an effect. From the model we see that on average, composite cognition scores tend to be lower compared to those who identify as white with the exception of native americans. So we we conclude race is infact associated with composite cognition. Lets look at race vs our exposure(threat and deprivation).

```{r}
boxplot(threat ~ race, data=midus)
boxplot(deprivation ~ race, data=midus)
```


### Sex

```{r}
t1 <-
midus %>%
  group_by(sex) %>% 
  summarize(
    mean_composite=mean(composite_cognition))

t2 <-
midus %>%
  group_by(sex) %>% 
  summarize(
    mean_episodic=mean(episodic_memory))

t3 <-
midus %>%
  group_by(sex) %>% 
  summarize(
    mean_exec=mean(executive_func))


knitr::kable(list(t1,t2,t3))
```

### Education


```{r}
boxplot(composite_cognition ~ education, data=midus)

education_lm <- lmer(composite_cognition ~ education + (1|family_id), data=midus)
summary(education_lm)
```


```{r}
boxplot(episodic_memory ~ education, data=midus)

education_lm2 <- lmer(episodic_memory ~ education + (1|family_id), data=midus)
summary(education_lm2)
```


```{r}
boxplot(executive_func ~ education, data=midus)

education_lm3 <- lmer(executive_func ~ education + (1|family_id), data=midus)
summary(education_lm3)
```

### Age


